---
title: 'Suicide Trend in Africa '
author: Daniel Boadzie
date: '2021-06-11'
slug: suicide-trend-in-africa
categories:
  - Data Analysis
tags:
  - Data Analysis
  - Data Cleaning
  - Analysis
description: ~
image: https://images.unsplash.com/photo-1564121211835-e88c852648ab?ixid=MnwxMjA3fDB8MHxwaG90by1wYWdlfHx8fGVufDB8fHx8&ixlib=rb-1.2.1&auto=format&fit=crop&w=1050&q=80
math: ~
license: ~
hidden: no
comments: yes
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load libraries for analysis
library(pacman)

p_load(tidyverse, ggpubr, ggthemes)
```



Suicide is a one of the major causes of death in our society today. Being a victim of suicide can be a traumatizing experience that is hard to forget. In this report we analyze the suicide trends across the world with emphasis on Ghana.

### Ask

1. Which year recorded the highest suicide?
2. What are the numbers for Ghana?
3. Why is it so?



### Prepare

The data is provided by World Bank via [Gapminder.org](https://www.gapminder.org/data/) under CC-BY license.

The has two columns country and a range of years from 1990 to 2016.


### Process

```{r}
# load the data
suicide <- read_csv("suicide_total_deaths.csv")


# get a glimpse
glimpse(suicide)
```

#### Data Formatting
We can see that we are dealing with a wide dataset that we need to make long for easy view and analysis.

```{r}
suicide <- suicide %>% 
  pivot_longer(`1990`:`2016`, names_to = "year", values_to = "deaths")

glimpse(suicide)
```

Now lets convert deaths column to an integer. But first we need to replace `k` with 1000.

```{r}
# replace `k` (1000) with actual values
suicide$deaths <- as.numeric(sub("k", "e3", suicide$deaths,  fixed = TRUE)) 

glimpse(suicide)

```

Lets convert deaths to numeric

```{r}
suicide <- suicide %>% 
  mutate(deaths = as.numeric(deaths))

glimpse(suicide)
```


#### Check for NAs

```{r}
colSums(is.na(suicide)) # check for NAs
```

Convert year column to numeric
```{r}
suicide %>% 
  mutate(year = as.integer(year))
```


### Analyze

#### Summary Statistics
Get summary statistics on the `deaths` column

```{r}
suicide$deaths %>% 
  summary()
```

Write new data to `csv` for further analysis.

```{r}
write_csv(suicide, "suicide_total_deaths_clean.csv")
```


#### Visualization

```{r, fig.align='center', fig.width=14, fig.height=10}
suicide %>% 
  select(country, deaths, year) %>% 
  filter(country == c("Ghana", "Nigeria", "Gambia") ) %>% 
  group_by(year) %>% 
  ggplot(aes(x=country, y=deaths)) +
  geom_col(aes(fill=year), position = "dodge", size=4) +
  coord_flip() +
  theme(legend.position = "bottom", plot.title = element_text(size = 20), axis.title.x = element_text(size = 18),  axis.title.y = element_text(size = 18)) +
   labs(title = "Suicide Deaths in Africa ", x = "Country", y="Deaths", fontface = "bold") +
  theme_economist()
  
```

The clearly an upward trend in suicide deaths in Africa. It is worth investigating why this might be the case.

In this article we cleaned, analyzed and visualized world suicide data with emphasis on Africa. 



