---
title: "Data Exploration with `DataExplorer`"
author: "Daniel Boadzie"
date: "2021-05-17"
categories: ["Data Cleaning"]
tags: ["Exporation", "Data Exploration", "EDA"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Introduction 

Getting to know your data is crucial in your analysis. That is why Exploratory Data Analysis is an important phase of the Data Analysis(EDA) process. After all, the better you understand your data, the easier it will be to extra meaningful insight from it. 

With the right tools, this all important phase can be a breeze especially when working in R and using `Dataexplorer` package.

## Data source

In this article we will learn to dig into data using `dataexplorer` from R. We will use the `diamonds` dataset available in Rstudio as our data source. The A dataset contains the prices and other attributes of almost 54,000 diamonds. It has the following attributes;

- price - price in US dollars (\$326–\$18,823)
- carat - weight of the diamond (0.2–5.01)
- cut   - quality of the cut (Fair, Good, Very Good, Premium, Ideal)
- color - diamond colour, from D (best) to J (worst)
- clarity - a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))
- x     - length in mm (0–10.74)
- y     - width in mm (0–58.9)
- z     - depth in mm (0–31.8)
- depth - total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43–79)
- table - width of top of diamond relative to widest point (43–95)


You may have to install the package `install.packages("Dataexplorer")` if you haven't done so.


```{r}
# load the packages
library(DataExplorer)
library(tidyverse)

# create our dataframe
df <- diamonds

# the data has the following columns
colnames(df)

# There are 53940 rows and 10 columns in the dataset
dim(df)
```

Now lets use `Dataexplorer` to examine our data.

```{r cache=TRUE}
introduce(df)
```

```{r}
plot_intro(df) #ploting the meta of the dataset
```

### Missing Values

Now lets plot the missing values in the data.

```{r}
plot_missing(df)

# lets confirm 
sum(is.na(df))
```

### Visualize the Distribution

```{r}
# we can visualize the distribution of the categorical data
plot_bar(df, ggtheme = theme_bw())
```

We can also create a chart based on some criteria.

```{r}
cut_filter <- df %>% 
  filter(., cut == c("Ideal", "Very Good")) %>% 
  plot_bar()
```

You can also plot a histogram. There are other charts type you can also explore.
 
```{r}
plot_histogram(df)
```
 
### Correlation Map

We can also create a correlation heatmap for all the features.
The degree of correlation is shown in the color legend with stronger correlation being red and weaker correlation being blue.

```{r }
plot_correlation(na.omit(df), maxcat = 5L)
```

### Create a Report

`Dataexplorer` can also help you create a report of your data. When you run the code below, you will get a beautiful and complete `HTML` document containing your report. It's just awesome!

```{r eval=FALSE }
create_report(df)
```


There is a lot more cool features of the `Dataexplorer` library that you can use when performing EDA. I suggest you take at this awesome package. You will be glad you did.
